---
title: "Random_Forest"
author: "Mohammad Mehralian"  
format:
  html:
    code-fold: true 
    embed-resources: true
    toc: true
    number-sections: true
    theme: cerulean
---
# Setup  

##Loading packages 
```{r setup, warning=FALSE, message=FALSE}
library(tidymodels)
library(tidyverse)
library(vip)
library(ranger)
library(finetune)
library(vip)
library(forcats)

```


```{r}
weather = read_csv("../data/merged_f.csv")%>%
  select(-DAP)

weather
read_csv("../data/merged_f.csv") %>% colnames()
```


# 1. Pre-processing
```{r }

set.seed(1681550) 

weather_split = initial_split(weather, 
  prop = .7, 
  strata = yield_mg_ha)

weather_split

```

## Data split

```{r }
weather_train = training(weather_split)  

weather_train 
```


```{r }
weather_test = testing(weather_split)    

weather_test
```


## Distribution of target variable

```{r }
ggplot() +
  geom_density(data = weather_train, 
               aes(x = yield_mg_ha),
               color = "red") +
  geom_density(data = weather_test, 
               aes(x = yield_mg_ha),
               color = "blue") 
  
```

## recipe
```{r }
weather_recipe <- recipe(yield_mg_ha ~ ., data = weather_train) %>%
  step_normalize(all_numeric_predictors())
  
weather_recipe
```

```{r }
weather_prep <- weather_recipe %>% 
  prep()

weather_prep
```

# 2. Training
## Model specification

```{r }
rf_spec <- rand_forest(
  trees = tune(),
  mtry = tune()
  ) %>%
  set_engine("ranger") %>%
  set_mode("regression")
rf_spec

```

## Cross-validation setup

We use 5-fold cross-validation to evaluate model performance during tuning:

```{r}

set.seed(1681550) 

resampling_foldcv = vfold_cv(weather_train, 
                              v = 4)

resampling_foldcv

resampling_foldcv$splits[[1]]

```

## Hyperparameter grid
```{r }
set.seed(1681550)
rf_grid_result <- tune_sim_anneal(object = rf_spec,
                     preprocessor = weather_recipe,
                     resamples = resampling_foldcv,
                    #param_info = rf_param,
                    iter = 4
                     )

beepr::beep()

rf_grid_result
rf_grid_result$.metrics[[2]]
```



```{r}
rf_grid_result %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  ggplot(aes(x = mtry, 
             y = trees 
             )) +
  geom_path(group = 1) +
  geom_point(aes(color = mean),
             size = 3) + 
  scale_color_viridis_b() +
  geom_text(aes(label = .iter), nudge_x = .0005) +
  labs(title = "RMSE")
```


```{r}
rf_grid_result %>%
  collect_metrics() %>%
  filter(.metric == "rsq") %>%
  ggplot(aes(x = mtry, 
             y = trees 
             )) +
  geom_path(group = 1) +
  geom_point(aes(color = mean),
             size = 3) + 
  scale_color_viridis_b() +
  geom_text(aes(label = .iter), nudge_x = .0005) +
  labs(title = "R2")
```


```{r }
best_rmse <- rf_grid_result %>%
  select_by_pct_loss("trees", 
                        metric = "rmse",
                        limit = 2
                     )
best_rmse


```

```{r}
best_r2 <- rf_grid_result  %>%
  select_by_pct_loss("trees", 
                        metric = "rsq",
                        limit = 2
                     )


best_r2
```

## final_spec
```{r}
final_spec <- rand_forest(trees = best_r2$trees,
                          mtry = best_r2$mtry) %>%
  # Specify the engine
  set_engine("ranger",
             importance = "permutation") %>%
    # Specifying mode  
  set_mode("regression")
  

final_spec
```



```{r}
final_fit <- last_fit(final_spec,
                weather_recipe,
                split = weather_split)

final_fit %>%
  collect_predictions()
```



```{r}
final_fit %>%
  collect_metrics()
```


## Metrics
```{r}
# RMSE
final_spec %>%
  fit(yield_mg_ha ~ .,
      data = bake(weather_prep, 
                  weather_train)) %>%
  augment(new_data = bake(weather_prep, 
                          weather_train)) %>% 
  rmse(yield_mg_ha, .pred) %>%
  bind_rows(
    
    
    # R2
    final_spec %>%
      fit(yield_mg_ha ~ .,
          data = bake(weather_prep, 
                      weather_train)) %>%
      augment(new_data = bake(weather_prep, 
                              weather_train)) %>% 
      rsq(yield_mg_ha, .pred)
    
  )
```


# 3.Predictions
```{r}
final_fit %>%
  collect_predictions() %>%
  ggplot(aes(x = yield_mg_ha,
             y = .pred)) +
  geom_point() +
  geom_abline() +
  geom_smooth(method = "lm") +
  scale_x_continuous(limits = c(0, 20)) +
  scale_y_continuous(limits = c(0, 20)) 
```

```{r}
library(ggplot2)


plot <- final_fit %>%
  collect_predictions() %>%
  ggplot(aes(x = yield_mg_ha, y = .pred)) +
  geom_point() +
  geom_abline() +
  geom_smooth(method = "lm") +
  scale_x_continuous(limits = c(0, 20)) +
  scale_y_continuous(limits = c(0, 20))


ggsave("predicted_vs_observed_plot.svg", plot = plot, device = "svg", width = 8, height = 6)


```


## variable importance
```{r}
final_spec %>%
  fit(yield_mg_ha ~ .,
         data = bake(weather_prep, weather)) %>%
    vi() %>%
  mutate(
    Variable = fct_reorder(Variable, 
                           Importance)
  ) %>%
  ggplot(aes(x = Importance, 
             y = Variable)) +
  geom_col() +
  scale_x_continuous(expand = c(0, 0)) +
  labs(y = NULL)
    
```




# 4. Prediction on Test data
```{r}
library(ggplot2)
library(dplyr)
library(forcats)


plot <- final_spec %>%
  fit(yield_mg_ha ~ .,
      data = bake(weather_prep, weather)) %>%
  vi() %>%
  mutate(
    Variable = fct_reorder(Variable, Importance)
  ) %>%
  ggplot(aes(x = Importance, 
             y = Variable)) +
  geom_col() +
  scale_x_continuous(expand = c(0, 0)) +
  labs(y = NULL)


ggsave("variable_importance_plot.svg", plot = plot, device = "svg", width = 8, height = 6)


```

```{r}
library(readr)
library(dplyr)

```

```{r }
weather_test <- read_csv("../data/testing/weather_2024A.csv") %>%
  distinct(year, site, .keep_all = TRUE)

trait_test <- read_csv("../data/testing/testing_submission.csv")

trait_weather_test <- left_join(trait_test, weather_test, by = c("year", "site"))

soil_test <- read_csv("../data/testing/testing_soil.csv") %>%
  distinct(year, site, .keep_all = TRUE)

tw_soil_test <- left_join(trait_weather_test, soil_test, by = c("year", "site"))

meta_test <- read_csv("../data/testing/testing_meta.csv")

tw_sm_meta_test <- left_join(tw_soil_test, meta_test, by = c("year", "site")) %>%
  mutate(
    yield_mg_ha = as.numeric(yield_mg_ha),
    swe_kg_m_2  = as.numeric(swe_kg_m_2)
  )

final_merged_test <- tw_sm_meta_test #%>%
  #select(-year, -hybrid, -previous_crop,
        # -longitude, -latitude, -lon, -lat)
#colnames(final_merged_test)

```


## Random Forest — Training and Prediction
```{r }
best_rf <- rf_grid_result %>%
  select_best(metric = "rsq")


final_spec <- rand_forest(trees = best_rf$trees,
                          mtry = best_rf$mtry) %>%
  set_engine("ranger", importance = "permutation") %>%
  set_mode("regression")

final_rf_wf <- workflow() %>%
  add_recipe(weather_recipe) %>%
  add_model(final_spec) %>%
  finalize_workflow(best_rf)

fitted_rf <- final_rf_wf %>%
  fit(data = training(weather_split))

predictions <- predict(fitted_rf, new_data = final_merged_test) %>%
  bind_cols(final_merged_test %>% select(site)) %>%
  rename(predicted_yield = .pred)

```

```{r}
dir.create("output", showWarnings = FALSE)
saveRDS(fitted_rf, file = "output/final_rf_model.rds")
```


```{r }
write_csv(predictions, "../data/testing/Predicted_RandomForest_test.csv")
```


```{r}
View(predictions)           
```

```{r}
library(ggplot2)

ggplot(predictions, aes(x = site, y = predicted_yield)) +
  geom_col() +
  theme_minimal() +
  labs(title = "Predicted Maize Yield by Site", x = "Site", y = "Predicted Yield (mg/ha)")

beepr::beep()
```













